# Vectors spaces {#vector-spaces}

```{r}
library(shiny)
```

## Linear equations

Let $x_1, x_2, \ldots, x_n$ be variables with coefficients $a_1, a_2, \ldots, a_n$, and $b$ are fixed and known numbers. Then, we say

$$
\begin{align}
\label{eq:linear}
a_1 x_1 + a_2 x_2 + \cdots + a_n x_n & = b
\end{align}
$$

is a linear equation. For example, the equation for a line with slope $m$ and $y$-intercept $b$ is

$$
\begin{align*}
y & = m x + b,
\end{align*}
$$

is a linear equation because it can be re-written as 

$$
\begin{align*}
y - m x & = b,
\end{align*}
$$

where $a_1 = 1$, $a_2 = m$, $x_1 = y$ and $x_2 = x$.

The equations

$$
\begin{align*}
\sqrt{19} x_1 & = (4 + \sqrt{2}) x_2 - x_3 - 9
& \mbox{ and } && 
-4 x_1 + 5 x_2 - 11 & = x_3
\end{align*}
$$

are both linear equations because they can be written as 

$$
\begin{align*}
\sqrt{19} x_1 - (4 + \sqrt{2}) x_2 + x_3 & = - 9 
& \mbox{ and } &&
-4 x_1 + 5 x_2 - x_3 & = 11,
\end{align*}
$$

respectively. The equations

$$
\begin{align*}
x_1 & = x_2^2 + 3
& \mbox{ and } && 
x_1 + x_2 - x_1 x_2 & = 16
\end{align*}
$$
are not linear equations because they do not meet the form of \ref{eq:linear} (The first equation above has a quadtric power of $x_2$ and the second equation has a product of $x_1$ and $x_2$).

### Systems of linear equations

A set of two or more linear equations that each contain the same set of variables is called a **system of linear equations**. The equations

$$
\begin{align*}
x_1 && + && 4 x_2 && - && x_3 && = & 11 \\
4 x_1 && + && 5 x_2 &&  && && = & 9
\end{align*}
$$

are a system of equations. Note that in the second equation, the coefficient for $x_3$ is 0, meaning we could re-write the above example as

$$
\begin{align*}
x_1 && + && 4 x_2 && - && x_3 && = & 11 \\
4 x_1 && + && 5 x_2 && + && 0 x_3 && = & 9.
\end{align*}
$$



## Vector spaces in $\mathcal{R}^n$

Let $\mathcal{R}^n$ be a real coordinate space of $n$ dimesions. You are already familiar with the Cartesian plane that consists of ordered pairs $(x, y)$. The Cartesian plane defines the real coordinate space $\mathbf{R}^2$ of two dimensions. In $\mathbf{R}^2$, the location of any point of interest can be defined using the $x$ and $y$. For example, the plot below shows the location of the point (2, 3)

```{r}
dat <- data.frame(
    x = 2,
    y = 3
)

ggplot(data = dat, aes(x = x, y = y)) +
    geom_point() +
    xlim(c(-4, 4)) +
    ylim(c(-4, 4))
```

A vector space is a generalization of this representation. In $\mathcal{R}^2$, we say that the vector $\mathbf{z} = c(2, 3)$ is centered at the origin (0, 0) and has length 2 in the $x$-axis and length 3 in the $y$-axis. The plot below shows this vector

```{r, echo = FALSE}
ggplot(data = dat, aes(x = x, y = y)) +
    geom_segment(aes(x = 0, xend = x, y = 0, yend = y), arrow = arrow()) +
    xlim(c(-4, 4)) +
    ylim(c(-4, 4))
```

We can also decompose the vector $\mathbf{z}$ into its $x$ and $y$ components. The $x$ component of $\mathbf{z}$ is (2, 0) and the $y$ component of $\mathbf{z}$ is (0, 3). The following plot shows the $x$ component (2, 0) in blue and the $y$ component (0, 3) in red.

```{r, echo = FALSE}
ggplot(data = dat, aes(x = x, y = y)) +
    geom_segment(aes(x = 0, xend = x, y = 0, yend = y), arrow = arrow()) +
    geom_segment(aes(x = 0, xend = x, y = 0, yend = 0), arrow = arrow(), color = "blue") +
    geom_segment(aes(x = 0, xend = 0, y = 0, yend = y), arrow = arrow(), color = "red") +
    xlim(c(-4, 4)) +
    ylim(c(-4, 4))
```


The below Shiny app allows you to plot the vector for any $(x, y)$ pair of your choosing.

```{r, echo = FALSE, eval = TRUE}
if (interactive()) {
  library(shiny)
  runGitHub(rep = "multivariable-math", username = "jtipton25", subdir = "shiny-apps/chapter-03/vector-space") 
}
if (!interactive()) {
  shinyApp(
    ui = fluidPage(
      sliderInput(
        inputId = "x",
        label = "x:",
        min = -10,
        max = 10,
        value = 2,
        step = 0.01
      ),
      sliderInput(
        inputId = "y",
        label = "y:",
        min = -10,
        max = 10,
        value = 2,
        step = 0.01
      ),
      plotOutput("Plot")
    ),
    
    server = function(input, output) {
      output$Plot = renderPlot({
        ggplot(data = data.frame(x = input$x, y = input$y),
               aes(x = x, y = y)) +
          geom_segment(aes(x = 0, xend = x, y = 0, yend = y), arrow = arrow()) +
          geom_segment(aes(x = 0, xend = x, y = 0, yend = 0), arrow = arrow(), color = "blue") +
          geom_segment(aes(x = 0, xend = 0, y = 0, yend = y), arrow = arrow(), color = "red") +
          xlim(c(-10, 10)) +
          ylim(c(-10, 10))
      })
    },
    
    options = list(height = 500)
  )
}
```

The shiny app can be downloaded and run on your own computer using 

```{r, eval = FALSE}
library(shiny)
runGitHub(rep = "multivariable-math", username = "jtipton25", subdir = "shiny-apps/chapter-03/vector-space") 
```

## Linear Combinations of Vectors

We say that for any two scalars $a$ and $b$ and any two vectors $\mathbf{x}$ and $\mathbf{y}$ of length $n$, the sum

$$
\begin{align*}
a \mathbf{x} + b \mathbf{y} & = \begin{pmatrix}
a x_1 + b y_1 \\
a x_2 + b y_2 \\
\vdots \\
a x_n + b y_n \\
\end{pmatrix}
\end{align*}
$$

is called a linear combination. The idea of a linear combination can be extended to $K$ different scalars $\{ a_1, \ldots, a_K \}$ and $K$ different vectors $\{ \mathbf{x}_1, \ldots, \mathbf{x}_K\}$ each of length $n$ as

$$
\begin{align*}
\sum_{k=1}^K a_k \mathbf{x}_k & = \begin{pmatrix}
\sum_{k=1}^K a_k x_{k1} \\
\sum_{k=1}^K a_k x_{k2} \\
\vdots \\
\sum_{k=1}^K a_k x_{kn} \\
\end{pmatrix}
\end{align*}
$$


