# Vectors spaces {#vector-spaces}


- [3 Blue 1 Brown -- Linear combinations, span, and basis vectors](https://www.3blue1brown.com/lessons/span)

```{r setup-02}
library(shiny)
library(patchwork)
library(tidyverse)
# if gg3D package not installed, install the package
library(gg3D)
library(dasc2594)
```

## Vectors 


### Properties of Vectors

For any real valued scalars $a, b \in \mathcal{R}$ and any vectors $\mathbf{x}, \mathbf{y}, \mathbf{z} \in \mathcal{R}^n$ (vectors of real numbers of length $n$), 

1) **scalar multiplication**
$$
\begin{aligned}
a \mathbf{x} & = a \begin{pmatrix} x_1 \\ x_2 \\ \vdots \\ x_n \end{pmatrix} \\
& = \begin{pmatrix} a x_1 \\ a x_2 \\ \vdots \\ a x_n \end{pmatrix} 
\end{aligned}
$$
    
where the scalar $a$ is multiplied by each element of the vector. For example, 

$$
\begin{aligned}
4 \begin{pmatrix} 4 \\ 6 \\ 7 \\ 12 \end{pmatrix} 
& = \begin{pmatrix} 4 * 4 \\ 4 * 6 \\ 4 * 7 \\ 4 * 12 \end{pmatrix} \\
& = \begin{pmatrix} 16 \\ 24 \\ 28 \\ 48 \end{pmatrix} 
\end{aligned}
$$

In `R`, we can multiply the vector by a scalar as

```{r scalar-multiply-vector}
4 * c(4, 6, 7, 12)
```

or if the vector $\mathbf{x} = \left( 4, 6, 7, 12 \right)'$ we can write this as

```{r scalar-multiply-vector2}
x <- c(4, 6, 7, 12)
4 * x
```

2) **scalar multiplicative commutivity**

$$
\begin{aligned}
a (b \mathbf{x}) & = (ab) \mathbf{x} & = b (a \mathbf{x})
\end{aligned}
$$
    
```{r scalar-multiply-commutivity}
4 * (6 * x) 
(4 * 6) * x
```

3) **scalar additive associativity**

$$
\begin{aligned}
a \mathbf{x} + b \mathbf{x} & = (a + b) \mathbf{x}
\end{aligned}
$$

4) **vector additive associativity**

$$
\begin{aligned}
a \mathbf{x} + a \mathbf{y} & = a (\mathbf{x} + \mathbf{y})
\end{aligned}
$$
    
5) **vector associativity**
$$
\begin{aligned}
\mathbf{x} + \mathbf{y} & = \mathbf{y} + \mathbf{x}
\end{aligned}
$$

$$
\begin{aligned}
(\mathbf{x} + \mathbf{y}) + \mathbf{z} & = \mathbf{x} + (\mathbf{y} + \mathbf{z})
\end{aligned}
$$


```{r add-vectors}
x <- c(1, 2, 3, 4)
y <- c(4, 3, 5, 1)
z <- c(5, 2, 4, 6)

x + y
y + x

(x + y) + z
x + (y + z)
```

6) **Identity Element of Addition:** For any vector $\mathbf{x}$ of length $n$, there exists a vector $\mathbf{0}$, known as the *zero vector*, such that

$$
\begin{aligned}
\mathbf{x} + \mathbf{0} & = \mathbf{x}
\end{aligned}
$$

```{r identity-add}
x + 0
x + rep(0, 4)
```

7) **Inverse Element of Addition:** For any vector $\mathbf{x}$ of length $n$, there exists a vector $-\mathbf{x}$, known as the *additive inverse* vector, such that

$$
\begin{aligned}
\mathbf{x} + (- \mathbf{x}) & = \mathbf{0}
\end{aligned}
$$

```{r add-inverse}
x + (-x)
```


    
## Vector addition
Two vectors of length $n$ can be added elementwise

$$
\begin{aligned}
\mathbf{x} + \mathbf{y} & = \begin{pmatrix} x_1 \\ x_2 \\ \vdots \\ x_n \end{pmatrix} + \begin{pmatrix} y_1 \\ y_2 \\ \vdots \\ y_n \end{pmatrix} \\
& = \begin{pmatrix} x_1 + y_1 \\ x_2 + y_2 \\ \vdots \\ x_n + y_n \end{pmatrix} 
\end{aligned}
$$

For example, 

$$
\begin{aligned}
\begin{pmatrix} 3 \\ 1 \\ -4 \\ 3 \end{pmatrix} + \begin{pmatrix} -3 \\ 17 \\ -39 \\ 4 \end{pmatrix} & = \begin{pmatrix} 3 + (-3) \\ 1 + 17 \\ -4 + (-39) \\ 3 + 4 \end{pmatrix} \\
& = \begin{pmatrix} 0 \\ 18 \\ -43 \\ 7 \end{pmatrix} 
\end{aligned}
$$

In `R`, we have

```{r vector-add-r}
x <- c(3, 1, -4, 3)
y <- c(-3, 17, -39, 4)
x + y
``` 

If two vectors $\mathbf{x}$ and $\mathbf{y}$ are of different lengths, then they cannot be added together. Using `R`, we get the following error:

```{r vector-add-r2}
x <- c(1, 2, 3)
y <- c(1, 2, 3, 4)
x + y
``` 

The error is telling us that the vector $\mathbf{x}$ and the vector $\mathbf{y}$ do not have the same length. 

Be careful when adding vectors in `R`. `R` uses "recycling" which means two vectors of different lengths can be added together if one vector is of a length that is a multiple of the other vector. For example, if $\mathbf{x} = (1, 2)'$ is a vector of length 2 and $\mathbf{y} = (1, 2, 3, 4)$ is a vector of length 4, `R` will add $\mathbf{x} + \mathbf{y}$ by replicating the vector $\mathbf{x}$ twice (i.e., $\mathbf{x} + \mathbf{y} = \left( \mathbf{x}', \mathbf{x}' \right)' = \left(1, 2, 1, 2 \right)' + \mathbf{y}$) 

```{r vector-add-recycle}
x <- c(1, 2)
y <- c(1, 2, 3, 4)
x + y
# replicated x = c(1, 2, 1, 2)
c(1, 2, 1, 2) + y
```

### The geometric interpretation of vectors in $\mathcal{R}^2$

Let $\mathcal{R}^2$ be a real coordinate space of $2$ dimensions. You are already familiar with the Cartesian plane that consists of ordered pairs $(x, y)$. The Cartesian plane defines the real coordinate space $\mathbf{R}^2$ of two dimensions. In $\mathbf{R}^2$, the location of any point of interest can be defined using the $x$ and $y$. For example, the plot below shows the location of the point (2, 3)

```{r geometric-interpretation}
dat <- data.frame(
    x = 2,
    y = 3
)

ggplot(data = dat, aes(x = x, y = y)) +
    geom_point() +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4))
```

A vector space is a generalization of this representation. In $\mathcal{R}^2$, we say that the vector $\mathbf{z} = c(2, 3)$ is centered at the origin (0, 0) and has length 2 in the $x$-axis and length 3 in the $y$-axis. The plot below shows this vector

```{r plot-vector, echo = FALSE}
ggplot(data = dat, aes(x = x, y = y)) +
    geom_segment(aes(x = 0, xend = x, y = 0, yend = y), arrow = arrow()) +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4))
```

We can also decompose the vector $\mathbf{z}$ into its $x$ and $y$ components. The $x$ component of $\mathbf{z}$ is (2, 0) and the $y$ component of $\mathbf{z}$ is (0, 3). The following plot shows the $x$ component (2, 0) in blue and the $y$ component (0, 3) in red.

```{r plot-vector-components, echo = FALSE}
ggplot(data = dat, aes(x = x, y = y)) +
    geom_segment(aes(x = 0, xend = x, y = 0, yend = y), arrow = arrow()) +
    geom_segment(aes(x = 0, xend = x, y = 0, yend = 0), arrow = arrow(), color = "blue") +
    geom_segment(aes(x = 0, xend = 0, y = 0, yend = y), arrow = arrow(), color = "red") +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4))
```


The below Shiny app allows you to plot the vector for any $(x, y)$ pair of your choosing.

```{r shiny-app-vectors, echo = FALSE, eval = FALSE}
library(shiny)
runGitHub(rep = "multivariable-math",
          username = "jtipton25", 
          subdir = "shiny-apps/chapter-03/vector-space") 
```

The shiny app can be downloaded and run on your own computer using 

```{r shin-app-vectors2, eval = FALSE}
library(shiny)
runGitHub(rep = "multivariable-math", 
          username = "jtipton25",
          subdir = "shiny-apps/chapter-03/vector-space") 
```

#### Addition of vectors

We can represent the addition of vectors geometrically as well. Consider the two vectors $\mathbf{u}$ = (3, 2) and $\mathbf{v}$ = (-2, 1) where $\mathbf{u} + \mathbf{v}$ = (1, 3).

```{r, plot-add-vectors}
data.frame(x = c(3, -2, 1), y = c(2, 1, 3), vector = c("u", "v", "u+v")) %>%
    ggplot() +
    geom_point(aes(x = x, y = y, color = vector)) +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4))
```

We can represent the sum using vectors by adding $\mathbf{u}$ first then adding $\mathbf{v}$ to $\mathbf{u}$ or by adding $\mathbf{v}$ first and then $\mathbf{u}$ to get

```{r plot-add-vectors-parallelogram}
df <- data.frame(x = c(0, 3, 1, -2), y = c(0, 2, 3, 1))
p1 <- ggplot() +
    geom_segment(aes(x = 0, xend = 3, y = 0, yend = 2), arrow = arrow(), color = "blue") +
    geom_segment(aes(x = 3, xend = 3 - 2, y = 2, yend = 2 + 1), arrow = arrow(), color = "red") +
    geom_segment(aes(x = 0, xend = 3 - 2, y = 0, yend = 2 + 1), arrow = arrow(), color = "black") +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4)) + 
    geom_polygon(data = df, aes(x = x, y = y), fill = "grey", alpha = 0.5) +
    ggtitle("u + v")
p2 <- ggplot() +
    geom_segment(aes(x = 0, xend = -2, y = 0, yend = 1), arrow = arrow(), color = "red") +
    geom_segment(aes(x = -2, xend = -2 + 3, y = 1, yend = 1 + 2), arrow = arrow(), color = "blue") +
    geom_segment(aes(x = 0, xend = 3 - 2, y = 0, yend = 2 + 1), arrow = arrow(), color = "black") +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4)) +
    geom_polygon(data = df, aes(x = x, y = y), fill = "grey", alpha = 0.5) +
    ggtitle("v + u")
p1 + p2
```

Notice that the sum of these vectors defines a parallelogram where the sum $\mathbf{u} + \mathbf{v}$ is the diagonal of the shaded parallelogram. This geometric interpretation will serve as a basis for interpreting vector equations in higher dimensions where typical visualization methods fail.

### Scalar multiplication of vectors


We can represent the multiplication of a vector by a scalar geometrically as well. Consider the vector $\mathbf{u}$ = (3, 2) and the scalars $a = 2$ and $b = -1$. Then, we can plot $\mathbf{u}$, $a\mathbf{u}$, and $b\mathbf{u}$.

```{r plot-scalar-multiplication}
data.frame(x = c(3, 2 * 3, -1 * 3), y = c(2, 2 * 2, -1 * 2), vector = c("u", "a*u", "b*u")) %>%
    ggplot() +
    geom_point(aes(x = x, y = y, color = vector)) +
    geom_segment(aes(x = 0, xend = x, y = 0, yend = y, color = vector), arrow = arrow(), alpha = 0.75) +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-6, 6), ylim = c(-6, 6))
```

In fact, if $a$ is allowed to take on any values, then the set of all possible values of $a \mathbf{u}$ for all values of $a$ defines an infinite line

```{r plot-line}
ggplot() +
    geom_abline(slope = 2/3, intercept = 0) +  
    geom_point(aes(x = 3, y = 2)) +
    geom_segment(aes(x = 0, xend = 3, y = 0, yend = 2), arrow = arrow(), color = "black") +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4)) 
```


### The geometric interpretation of vectors in $\mathcal{R}^3$

Let the vector $\mathbf{u} = c(-2, 3, 5)$. Then, the figure below shows the vector in 3 dimensions.

**Draw picture by hand**
```{r plot-3d-vector, echo = FALSE}
# # uses gg3D library
# u <- c(-2, 3, 5)
# theta <- NULL
# phi <- NULL
# data.frame(x = u[1], y = u[2], z = u[3]) %>%
#     ggplot( aes(x = x, y = y, z = z, group = 1)) +
#     axes_3D(theta = theta, phi = phi) +
#     stat_3D(aes(x = x, y = y, z = z), geom = "line", theta = theta, phi = phi) +
#     theme_void() +
#     labs_3D(hjust=c(0,1,1), vjust=c(1, 1, -0.2), angle=c(0, 0, 90), theta = theta, phi = phi) 
# data.frame(x = u[1], y = u[2], z = u[3]) %>%
#     ggplot( aes(x = x, y = y, z = z, group = 1)) +
#     axes_3D() +
#     stat_3D(aes(x = x, y = y, z = z, group = "1"), geom = "line") +
#     theme_void() +
#     labs_3D(hjust=c(0,1,1), vjust=c(1, 1, -0.2), angle=c(0, 0, 90))
```

### The geometric interpretation of vectors in $\mathcal{R}^n$

As the number of dimensions increases, the same interpretation can be used, but the ability to visualize higher dimensions becomes more difficult.


### Linear Combinations of Vectors

We say that for any two scalars $a$ and $b$ and any two vectors $\mathbf{x}$ and $\mathbf{y}$ of length $n$, the sum

$$
\begin{aligned}
a \mathbf{x} + b \mathbf{y} & = \begin{pmatrix}
a x_1 + b y_1 \\
a x_2 + b y_2 \\
\vdots \\
a x_n + b y_n \\
\end{pmatrix}
\end{aligned}
$$

is called a linear combination. The idea of a linear combination can be extended to $K$ different scalars $\{ a_1, \ldots, a_K \}$ and $K$ different vectors $\{ \mathbf{x}_1, \ldots, \mathbf{x}_K\}$ each of length $n$ as

$$
 \begin{aligned}
a_1 \mathbf{x}_1 + a_2 \mathbf{x}_2 +  \ldots + a_K \mathbf{x}_K = 
\sum_{k=1}^K a_k \mathbf{x}_k & = \begin{pmatrix}
\sum_{k=1}^K a_k x_{k1} \\
\sum_{k=1}^K a_k x_{k2} \\
\vdots \\
\sum_{k=1}^K a_k x_{kn} \\
\end{pmatrix}
\end{aligned}
$$
The scalars $a_k$ are called **coefficients** (sometimes also called **weights**). 

* **Example:** Consider the linear combination $a \mathbf{u} + b \mathbf{v}$ where 

$$
\begin{aligned}
\mathbf{u} = \begin{pmatrix} 3 \\ 6\end{pmatrix} && \mathbf{v} = \begin{pmatrix} -2 \\ 1\end{pmatrix}. \end{aligned}
$$
Are there values of $a$ and $b$ such $a \mathbf{u} + b \mathbf{v} = \begin{pmatrix} 9 \\ - 4 \end{pmatrix}$? To answer this question, we can write the linear combination as 
$$
\begin{aligned}
a \begin{pmatrix} 3 \\ 6\end{pmatrix} + b \begin{pmatrix} -2 \\ 1\end{pmatrix} & = \begin{pmatrix} 9 \\ -4 \end{pmatrix} 
\end{aligned}
$$
which can be written using the property of scalar multiplication as
$$
\begin{aligned}
\begin{pmatrix} 3a \\ 6a \end{pmatrix} + \begin{pmatrix} -2b \\ b \end{pmatrix} & = \begin{pmatrix} 9 \\ -4 \end{pmatrix}
\end{aligned}
$$
and using properties of vector addition can be written as
$$
\begin{aligned}
\begin{pmatrix} 3a - 2b \\ 6a + b \end{pmatrix} & = \begin{pmatrix} 9 \\ -4 \end{pmatrix}
\end{aligned}
$$
Recognizing this as a system of linear equations
$$
\begin{aligned}
3a - 2b & = 9 \\ 
6a + b & = -4,
\end{aligned}
$$
the system of equations can be written in an augmented matrix form as
$$
\begin{aligned}
\begin{pmatrix}
3  & - 2 & 9\\
6  &   1 & -4 
\end{pmatrix}
\end{aligned}
$$
Reducing the augmented matrix to reduced row echelon form gives
```{r rref-augmented-matrix}
rref(matrix(c(3, 6, -2, 1, 9, -4), 2, 3))
```
which has solutions $a = 0.0667$ and $b = -4.4$.

* **Result:** Any vector equation $a_1 \mathbf{x}_1 + a_2 \mathbf{x}_2 +  \ldots + a_K \mathbf{x}_K = \mathbf{c}$ for a given constant vector $\mathbf{b}$ has the same solution set as the augmented matrix

$$
\begin{aligned}
\begin{pmatrix} \mathbf{x}_1 & \mathbf{x}_2 & \cdots & \mathbf{x}_K & \mathbf{b} \end{pmatrix}
\end{aligned}
$$
Equivalently, the set of vectors $\{\mathbf{x}_k\}_{k=1}^K$ can only be combined with linear coefficients $\{a_k\}_{k=1}^K$ to equal the vector $\mathbf{b}$ if the linear system of equations is consistent.

### The geometric interpretation of linear combinations of vectors

Consider the vectors $\mathbf{u} = \begin{pmatrix} \sqrt{2} \\ - \sqrt{2} \end{pmatrix}$ and $\mathbf{v} = \begin{pmatrix} 1 \\ 1 \end{pmatrix}$ shown in the figure below on the left.

```{r plot-linear-combination, echo = FALSE}
df <- data.frame(x = c(sqrt(2), 1), y = c(-sqrt(2), 1))
p1 <- ggplot() +
    geom_segment(aes(x = 0, xend = sqrt(2), y = 0, yend = -sqrt(2)), arrow = arrow(length = unit(0.1, "inches")), size = 1.5, color = "blue") +
    geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), arrow = arrow(length = unit(0.1, "inches")), size = 1.5, color = "red") +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4)) + 
    ggtitle("u and v")

transformation_matrix <- tribble(~ x, ~ y,
                                 2, 1,
                                 -2, 1) %>% 
  as.matrix()

grid_start <- construct_grid() %>% 
  mutate(id = row_number())

grid_trans <- grid_start %>% 
  # need to `transform_df_coords()` twice as each segment is made up of 2 points
  transform_df_coords(x, y, m = transformation_matrix) %>% 
  transform_df_coords(xend, yend, m = transformation_matrix)

grid_all <- bind_rows(
  mutate(grid_start, time = 1),
  mutate(grid_trans, time = 2)
)

basis_start <- tibble(
  x = c(0, 0),
  y = c(0, 0),
  xend = c(1, 0),
  yend = c(0, 1),
  # `vec` is unnecessary, will just use to differentiate colors
  vec = c("i", "j")
) %>% 
  mutate(id = nrow(grid_start) + row_number())

basis_trans <- basis_start %>% 
  transform_df_coords(x, y, m = transformation_matrix) %>% 
  transform_df_coords(xend, yend, m = transformation_matrix)

basis_all <- bind_rows(
  mutate(basis_start, time = 1),
  mutate(basis_trans, time = 2)
)

x_breaks <- unique(grid_start$x)
y_breaks <- unique(grid_start$y)


p2 <- ggplot(aes(x = x, y = y, group = id), data = filter(grid_all, time == 2))+
    geom_segment(aes(xend = xend, yend = yend)) +
    geom_segment(aes(xend = xend, yend = yend, color = vec), data = filter(basis_all, time == 2), arrow = arrow(length = unit(0.1, "inches")), size = 1.5) +
    scale_color_manual(values = c("i" = "blue", "j" = "red")) +
    scale_x_continuous(breaks = x_breaks, minor_breaks = NULL) +
    scale_y_continuous(breaks = y_breaks, minor_breaks = NULL) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4)) + 
    theme(legend.position = "none") +
    ggtitle("integer linear combinations \nof u and v")
p1 + p2
```



* **Exercise:** Given $\mathbf{u} = \begin{pmatrix} 3 \\ -1 \end{pmatrix}$ and $\mathbf{v} = \begin{pmatrix} 0 \\ 1 \end{pmatrix}$, estimate the linear combination of $\mathbf{u}$ and $\mathbf{v}$ that gives the point $\mathbf{w}$ in the figure below.


```{r plot-transformation, echo = FALSE}
transformation_matrix <- tribble(~ x, ~ y,
                                 3, 0,
                                 -1, 1) %>% 
  as.matrix()

grid_start <- construct_grid() %>% 
  mutate(id = row_number())

grid_trans <- grid_start %>% 
  # need to `transform_df_coords()` twice as each segment is made up of 2 points
  transform_df_coords(x, y, m = transformation_matrix) %>% 
  transform_df_coords(xend, yend, m = transformation_matrix)

grid_all <- bind_rows(
  mutate(grid_start, time = 1),
  mutate(grid_trans, time = 2)
)

basis_start <- tibble(
  x = c(0, 0),
  y = c(0, 0),
  xend = c(1, 0),
  yend = c(0, 1),
  # `vec` is unnecessary, will just use to differentiate colors
  vec = c("i", "j")
) %>% 
  mutate(id = nrow(grid_start) + row_number())

basis_trans <- basis_start %>% 
  transform_df_coords(x, y, m = transformation_matrix) %>% 
  transform_df_coords(xend, yend, m = transformation_matrix)

basis_all <- bind_rows(
  mutate(basis_start, time = 1),
  mutate(basis_trans, time = 2)
)

x_breaks <- unique(grid_start$x)
y_breaks <- unique(grid_start$y)


ggplot(aes(x = x, y = y, group = id), data = filter(grid_all, time == 2))+
    geom_segment(aes(xend = xend, yend = yend)) +
    geom_segment(aes(xend = xend, yend = yend, color = vec), data = filter(basis_all, time == 2), arrow = arrow(length = unit(0.1, "inches")), size = 1.5) +
    geom_point(aes(x = -3, y = 3), size = 2, color = "orange") +
    geom_text(data = data.frame(x = c(3, 0, -3), y = c(-1, 1, 3), text = c("u", "v", "w")),
              aes(x = x + 0.25, y = y + 0.25, label = text), size = 5, inherit.aes = FALSE,
              color = c("blue", "red", "orange")) +
    # geom_text(data = data.frame(x = -3, y = 3, text = "w"),
    #           aes(x = x, y = y, label = text), inherit.aes = FALSE) +
    scale_color_manual(values = c("i" = "blue", "j" = "red")) +
    scale_x_continuous(breaks = x_breaks, minor_breaks = NULL) +
    scale_y_continuous(breaks = y_breaks, minor_breaks = NULL) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4)) + 
    theme(legend.position = "none") +
    ggtitle("integer linear combinations \nof u and v")
```

## Span

:::{.definition #span name="Span"}
Let $\mathbf{a}_1, \ldots, \mathbf{a}_K$ be vectors in $\mathcal{R}^n$. We say the vector $\mathbf{w}$ is in the span of $\mathbf{a}_1, \ldots, \mathbf{a}_K$ ($\mathbf{w} \in \mbox{span}\{ \mathbf{a}_1, \ldots, \mathbf{a}_K \}$) if there exists coefficients $x_1, \ldots, x_K$ such that $\mathbf{w} = \sum_{k=1}^K x_k \mathbf{a}_k$.
:::

:::{.example}
While not a vector notation, you already understand the span from polynomial functions. For example, assume you have the functions 1, $x$, and $x^2$. Then, the functions $-4 + 3x^2$ ($a_1 = -4, a_2 = 0, a_3 = 3$) and $-3 + 4x - 2x^2$ ($a_1 = -3, a_2 = 4, a_3 = -2$) are in the span of functions $\{1, x, x^2\}$, but the functions $x^3$, $x^4 - 2x^2$, etc., are not in the span of $\{1, x, x^2\}$ because you cannot write these as a linear combination of $a_1 1 + a_2 x + a_3 x^2$. 
:::

### Geometric example of the span

* **Example:** Consider the vector $\mathbf{u} = \begin{pmatrix} 2 \\ 1 \end{pmatrix}$. Then, the vector $\mathbf{w} = \begin{pmatrix} 4 \\ 2 \end{pmatrix}$ is in the $\mbox{span}\{\mathbf{u}\}$ because $\mathbf{w} = 2 \mathbf{u}$ but the vector $\mathbf{v} = \begin{pmatrix} 4 \\ -4 \end{pmatrix}$ is not in the $\mbox{span}\{\mathbf{u}\}$ because there is no coefficient $a$ such that $\mathbf{w} = a \mathbf{u}$. In this example, the vector $\mathbf{u}$ is a 2-dimensional vector (lives in $\mathcal{R}^2$--a plane) but the $\mbox{span}\{\mathbf{u}\}$ lives in 1-dimension (a line).


```{r plot-example-span}
ggplot() +
    geom_abline(slope = 1/2, intercept = 0, color = "blue", size = 2) +  
    geom_segment(aes(x = 0, xend = 2, y = 0, yend = 1), arrow = arrow(length = unit(0.1, "inches")), size = 1.5, color = "red") +
    geom_segment(aes(x = 0, xend = 4, y = 0, yend = 2), arrow = arrow(length = unit(0.1, "inches")), size = 1.5, color = "red") +
    geom_segment(aes(x = 0, xend = 4, y = 0, yend = -4), arrow = arrow(length = unit(0.1, "inches")), size = 1.5, color = "orange") +
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4))  +
    geom_text(data = data.frame(x = c(2, 4, 4),  y = c(1, 2, -4), text = c("u", "w", "v")),
              aes(x = x, y = y + 0.5, label = text), size = 5, inherit.aes = FALSE,
              color = c("red", "red", "orange")) +
    ggtitle("span{u} is the blue line \nw is in span{u}\nv is not in span{u}")
```

From the example above, we can answer the question "Is the point (a, b) on the line defined by the vector $\mathbf{u}$?" by asking whether the point (a, b) is in the $span\{\mathbf{u}\}$. While this is trivial for such a simple problem, the use of the span will make things easier in higher dimensions.

* **Example: do in class** 2 3-d vectors that are not scalar multiples of each other define a plane. Does a point lie within the plane? Use the span to answer this question.


The shiny app below demonstrates how the concept of span can be understood in 2 dimensions. The app can be downloaded and run

```{r shiny-app-span, eval = FALSE}
library(shiny)
runGitHub(rep = "multivariable-math",
          username = "jtipton25", 
          subdir = "shiny-apps/chapter-03/span") 
```


