# Determinants {#determinants}

```{r, message = FALSE}
library(tidyverse)
library(dasc2594)
```

The determinant $\operatorname{det}(\mathbf{A})$ of a square $n \times n$ matrix $\mathbf{A}$ is a real number. 
```{definition, determinant}
The determinant is a fuction of a square $n \times n$ matrix that satisfies the following properties

a) If a scalar multiple of one row of $\mathbf{A}$ is added to another row of $\mathbf{A}$, then the determinant $\operatorname{det}(\mathbf{A})$ is unchanged.

b) Scaling a row of $\mathbf{A}$ by a constant $c$ multiples the determinant by $c$

c) Swapping two rows of a matrix $\mathbf{A}$ multiplies the determinant by -1

d) The determinant of the $n \times n$ identity matrix $\mathbf{I}$ is equal to 1
```

The determinant is the unique function mapping square matrices to the real number line that satisfies the above definition.

::: {.example}
Let $\mathbf{A} = \begin{pmatrix} 5 & 3 \\ 1 & -3 \end{pmatrix}$. Then, we can calculate the determinant $\operatorname{det}(\mathbf{A})$ using row operations to get to reduced row echelon form.

\begin{align*}
&& \begin{pmatrix} 5 & 3 \\ 1 & -3 \end{pmatrix} && \operatorname{det} = x \\
\mbox{swap row 1 and row 2} && \begin{pmatrix} 1 & -3 \\ 5 & 3  \end{pmatrix} && \operatorname{det} = -x \\
\mbox{row 2 = -5 * row 1 + row 2} && \begin{pmatrix} 1 & -3 \\ 0 & 18  \end{pmatrix} && \operatorname{det} = -x \\
\mbox{row 2 \div 18} && \begin{pmatrix} 1 & -3 \\ 0 & 1  \end{pmatrix} && \operatorname{det} = -\frac{x}{18} \\
\mbox{row 1 = row 1 + 3 * row 2} && \begin{pmatrix} 1 & 0 \\ 0 & 1  \end{pmatrix} && \operatorname{det} = -\frac{x}{18} 
\end{align*}
where the last matrix is in reduced row echelon form and is the identity matrix which has determinant $\operatorname{det}(\mathbf{I}) = 1$. Therefore $-\frac{x}{18} = 1$ which implies that $x = -18$ so that $\operatorname{det}(\mathbf{A}) = -18$
Let's check our answer in R

````{r}
A <- matrix(c(5, 1, 3, -3), 2, 2)
det(A)
````
:::

::: {.example}
In class
:::

## Determinants of $2 \times 2$ matrices

::: {.definition}
If $\mathbf{A} = \begin{pmatrix} a & b \\ c & d \end{pmatrix}$ is a $2 \times 2$ matrix, the determinant $\operatorname{det}(\mathbf{A}) = ad - bc$ 
:::
```{r, echo = FALSE}
A <- matrix(c(5, 1, 3, -3), 2, 2)
```

::: {.example}
Let $\mathbf{A} = `r array_to_latex(A)`$, then $\det(\mathbf{A}) = (5 * -3) - (3 * 1) = -18$
:::


::: {.example}
Let $\mathbf{A} = \begin{pmatrix} a & b \\ c & d \end{pmatrix}$. Then

a) If $a$ = 0

b) $a \neq 0$

show steps in class
:::

## Determinants of $n \times n$ matrices

To better understand determinants of $n \times n$ matrices, we need to define the two concepts of a matrix **minor** and **cofactor**.

::: {.definition}
For an $n \times n$ matrix $\mathbf{A}$, 

a) The (i, j) **minor** $\mathbf{A}_{-i-j}$ is the $(n-1) \times (n-1)$ matrix obtained by deleting the $i$th row and the $j$ column from $\mathbf{A}$

b) The (i, j) **cofactor** $c_{ij}$ is defined using the determinant of the  minor where 
\begin{align*}
c_{ij} = \mathbf(-1)^{i + j} \det{\mathbf{A}_{-i-j}}
\end{align*}
:::

Note: The cofactor of a scalar $a$ (a $1 \times 1$ matrix) is defined as $\mathbf{C}_{ij} = (-1)^{1 + 1} \det(a) = a$.

Note: The leading term in the cofactor definition $\mathbf(-1)^{i + j}$ defines a checkerboard pattern shown below
\begin{align*}
\begin{pmatrix}
+ & - & + & - & \cdots \\
- & + & - & + & \cdots \\
+ & - & + & - & \cdots \\
- & + & - & + & \cdots \\
\vdots & \vdots & \vdots & \vdots & \ddots \\
+ & - & + & - & \cdots \\
\end{pmatrix}
\end{align*}


::: {.example}
Let $\mathbf{A}$ be the $3 \times 3$ matrix... (in class)

a) Find the minor $\mathbf{A}_{-2-3}$

b) Find the cofactor $c_{23}$
:::

Note that in the cofactor definition of a $n \times n$ matrix it is assumed that you can calculate the determinant of the $n-1 \times n-1$ minor $\mathbf{A}_{-i-j}$. From this we see that each of the $n \times n$ cofactors of $\mathbf{A}$ are themselves the (signed) determinants of $n-1 \times n-1$ submatrices (the matrix minors). Thus, solving for all cofactors in general requires a recursive definition where smaller and smaller submatrices are evaluated. 

```{theorem, name = "Cofactor exapansion"}

Let $\mathbf{A}$ be an $n \times n$ matrix with $ij$th elements $a_{ij}$. Then

a) The **cofactor expansion along the $i$th row** (for any fixed row $i$) is 
\begin{align*}
\det(\mathbf{A}) = \sum_{j=1}^n a_{ij} c_{ij} = a_{i1} c_{i1} + a_{i2} c_{i2} + \cdots + a_{in} c_{in}
\end{align*}

b) The **cofactor expansion along the $j$th column** (for any fixed column $j$) is 
\begin{align*}
\det(\mathbf{A}) = \sum_{i=1}^n a_{ij} c_{ij} = a_{1j} c_{1j} + a_{2j} c_{2j} + \cdots + a_{nj} c_{nj}
\end{align*}

```


::: {.proof}
This is quite complex. For those interested, an example is available [here](https://textbooks.math.gatech.edu/ila/determinants-cofactors.html)
:::


::: {.note}
The above theorem states that there are actually $2n$ ways to calcuate the determinant--one for each row and column of $\mathbf{A}$.
:::

::: {.example}
in class: Use the minor/cofactor defintion to calculate the determinant of tha $2 \times 2$ matrix
:::


::: {.example}
in class: Use the minor/cofactor defintion to calculate the determinant of tha $3 \times 3$ matrix
:::


::: {.example}
in class: determinant of a matrix with a row (or column) of zeros
:::


::: {.example}
in class: determinant of a matrix where at least one row is a linear combination of the other rows
:::

::: {.theorem}
in class: determinant of a matrix is equal to the determinant of its transpose
:::

::: {.proof}
Follows directly from cofactor expansion theorem
:::


## Properties of determinants

::: {.theorem}
A $n \times n$ square matrix $\mathbf{A}$ is invertible if and only if $\det(\mathbf{A}) \neq 0$
:::

::: {.proof}
From the invertible matrix theorem **link to this in notes**, we know that the matrix $\mathbf{A}$ is invertible if and only if every column of $\mathbf{A}$ is a pivot column. Therefore, each column is linearly independent from the other columns. Based on the example above, if the rows were not linearly independent, the determinant would be equal to 0 (as row operations could create a row/column of all zeros). Thus, if the determinant is not 0, the columns of $\mathbf{A}$ are linearly independent and the matrix is invertible.
:::

::: {.theorem}
If $\mathbf{A}$ and $\mathbf{B}$ are $n \times n$ matrices, $\mathbf{I}$ is an $n \times n$ identity matrix, and $c$ is a scalar, we have

a) $\det(\mathbf{I}) = 1$

b) $\det(\mathbf{A}) = \det(\mathbf{A}')$

c) $\det(\mathbf{A}^{-1}) = 1 / \det(\mathbf{A})$ if $\det(\mathbf{A}) \neq 0$ ($\mathbf{A}$ is invertible)

d) $\det(\mathbf{A}\mathbf{B}) = \det(\mathbf{A})\det(\mathbf{B})$

e) $\det(c\mathbf{A}) = c^n \det(\mathbf{A})$

:::

## Cramer's Rule and Determinants

While commonly used for theoretical results, Cramer's rule is not commonly used in applied linear algebra. As such, we will mention Cramer's rule but not focus on it.

```{theorem, name = "Cramer's Rule"}
Let $\mathbf{A}$ be a $n \times n$ invertible matrix. Define $\mathbf{A}_i(\mathbf{b})$ as the matrix $\mathbf{A}$ with the $i$th column replace by the vector $\mathbf{b}$. For example, $\mathbf{A}_i(\mathbf{b}) = \begin{pmatrix} \mathbf{a}_1 & \cdots & \mathbf{a}_{i-1} & \mathbf{b} & \mathbf{a}_{i+1} & \cdots & \mathbf{a}_n \end{pmatrix}$. Then, for any $\mathbf{b} \in \mathcal{R}^n$, the unique solution to $\mathbf{A} \mathbf{x} = \mathbf{b}$ has entries given by 
\begin{align*}
x_i = \frac{\det(\mathbf{A}_i(\mathbf{b}))}{\det(\mathbf{A})} & \mbox{ for } i = 1, 2, \ldots, n
\end{align*}
```

```{example}
In Cramer's rule, why do we know

* the solution is unique for any $\mathbf{b}$? 

* the determinant $\det(\mathbf{A}) \neq 0$?
```
