# Linearly independent sets and bases

```{r, message = FALSE}
library(tidyverse)
library(dasc2594)
```

Recall that a set of vectors $\{ \mathbf{v}_1, \ldots, \mathbf{v}_n\}$ is linearly independent if the only solution to the system of equations 

\begin{align*}
x_1 \mathbf{v}_1 + \cdots + x_n \mathbf{v}_n = \mathbf{0}
\end{align*}
is the trivial solution $\mathbf{x} = \mathbf{0}$. In other words, it is not possible to write any of the vectors in the set $\{ \mathbf{v}_1, \ldots, \mathbf{v}_n\}$ as a linear combination of the other vectors. 


:::{.definition}
Let $\mathcal{H}$ be a subspace of a vector space $\mathcal{V}$. Then, the set of vectors $\mathcal{B} = \{ \mathbf{v}_1, \ldots, \mathbf{v}_n \}$ is a **basis** for $\mathcal{H}$ if

* The set of vectors $\mathcal{B}$ are linearly independent

* The subspace spanned by $\mathcal{B}$ is $\mathcal{H}$. In other words
\begin{align*}
\mbox{span}(\mathbf{v}_1, \ldots, \mathbf{v}_n) = \mathcal{H}
\end{align*}
:::


:::{.example}
* in class--standard basis $\mathbf{e}_1, \ldots \mathbf{e}_n$ which are the columns of the $n \times n$ identity matrix$\mathbf{I}$.
:::

:::{.example}
* in class--pick 3 vectors of length 3. Are they a basis for $\mathcal{R}^3$? What about $\mathcal{R}^4$?
:::


```{theorem, name = "The Spanning Set Theorem"}
Let $\mathcal{S} = \{\mathbf{v}_1, \ldots, \mathbf{v}_n\}$ be vectors in the vector space $\mathcal{V}$ and let $\mathcal{H}$ = span($\mathbf{v}_1, \ldots, \mathbf{v}_n$)

a) If one of the vectors, say $\mathbf{v}_k$, of $\mathcal{S}$ is a linear combination of the remaining vectors of $\mathcal{S}$, then the set formed by removing the vector $\mathbf{v}_k$ still spans $\mathcal{H}$
    
b) If $\mathcal{H} \neq \{\mathbf{0}\}$, then some subset of $\mathcal{S}$ spans $\mathcal{H}$.
```

```{proof}
a) WLOG assume $\mathbf{v}_n$ is a linear combination of $\mathbf{v}_1, \ldots, \mathbf{v}_{n-1}$ (if not, permute the labels to make the linearly dependent vector the $n$th vector). Then
\begin{align}
(\#eq:span-set-theorem1)
\mathbf{v}_n = x_1 \mathbf{v}_1 + \cdots + x_{n-1} \mathbf{v}_{n-1}
\end{align}
Because the vectors $\{\mathbf{v}_1, \ldots, \mathbf{v}_n\}$ span $\mathcal{H}$, any vector $\mathbf{b}\in \mathcal{H}$ can be written as 
\begin{align}
(\#eq:span-set-theorem2)
\mathbf{b} = c_1 \mathbf{v}_1 + \cdots + c_{n} \mathbf{v}_{n}
\end{align}
for scalars $c_1, \ldots, c_n$. Plugging the result from \@ref(eq:span-set-theorem1) into \@ref(eq:span-set-theorem2) shows that any vector $\mathbf{b}$ in $\mathcal{H}$ can be written only using the vectors $\mathbf{v}_1, \ldots, \mathbf{v}_{n-1}$

b) As the vectors in $\mathcal{S}$ span $\mathcal{H}$, if there is a linearly dependent vector in $\mathcal{S}$, this vector can be removed from $\mathcal{S}$ and the span of this subset of $\mathcal{S}$ will still span $\mathcal{H}$. As long as $\mathbf{H} \neq \{\mathbf{0}\}$, there must be a least one nonzero vector in $\mathbf{S}$ so the removing of linearly dependent vectors will stop with at least one vector. As all of the linearly dependent vectors have been removed, the subset of $\mathcal{S}$ created in this manner will be a set of linearly independent vectors that span $\mathcal{H}$.
```

## Bases for null($\mathbf{A}$) and col($\mathbf{A}$)

:::{.example}
Find a basis for the null($\mathbf{A}$) where
```{r}
set.seed(2021)
A <- matrix(sample(-9:9, 15, replace = TRUE), 3, 5)
```
\begin{align*}
\mathbf{A} = `r array_to_latex(A)`
\end{align*}

:::

