# Change of basis

- [3 Blue 1 Brown -- Change of basis](https://www.3blue1brown.com/lessons/change-of-basis)

```{r setup-18}
library(tidyverse)
library(dasc2594)
```

Consider two bases $\mathcal{B} = \{ \mathbf{b}_1, \ldots, \mathbf{b}_n \}$ and $\mathcal{C} = \{ \mathbf{c}_1, \ldots, \mathbf{c}_n \}$ for a vector space $\mathcal{V}$. If we have a vector $\left[\mathbf{x}\right]_B$ with coordinates in $\mathcal{B}$, what are the coordinates of $\left[\mathbf{x}\right]_C$ with respect to $\mathcal{C}$?

* Recall: we know how to change from the standard coordinates to the basis $\mathcal{B}$. If $\mathbf{x}$ is a vector in the standard coordinates and $\mathbf{B} = \begin{pmatrix} \mathbf{b}_1 & \ldots & \mathbf{b}_n \end{pmatrix}$ is a matrix with columns given by the basis $\mathbf{B}$, the coordinates of $\left[\mathbf{x}\right]_B$ of the vector $\mathbf{x}$ with respect to the basis $\mathcal{B}$ are 
$$
\begin{aligned}
\left[\mathbf{x}\right]_B = \mathbf{B}^{-1} \mathbf{x}
\end{aligned}
$$
and, as a consequence, given a vector $\left[\mathbf{x}\right]_B$ with coordinates with respect to the basis $\mathcal{B}$, the vector of coefficients $\mathbf{x}$ with standard coordinates is given by 
$$
\begin{aligned}
\mathbf{x} = \mathbf{B} \left[\mathbf{x}\right]_B.
\end{aligned}
$$

Notice that change of coordinates is a linear transformation from $\mathcal{B}$ to $\mathcal{C}$ with transformation matrix $\mathbf{A}$. Despite the more complex notation, this is just another linear transformation [[link](https://www.youtube.com/watch?v=VGhij2qmOs4)].

::: {#exm-}
Show the change of basis from the basis $\mathcal{B} = \left\{ \begin{pmatrix} \frac{1}{2} \\ 1 \end{pmatrix}, \begin{pmatrix} -1 \\ 0 \end{pmatrix}\right\}$ to the basis $\mathcal{C} = \left\{ \begin{pmatrix} 0 \\ \frac{1}{2} \end{pmatrix}, \begin{pmatrix} 1 \\ -\frac{1}{2} \end{pmatrix}\right\}$. To do this, represent the columns that make up the basis $\mathcal{B}$ as the matrix $\mathbf{B} = \begin{pmatrix} \frac{1}{2} & -1 \\ 1 & 0 \end{pmatrix}$ and represent the columns that make up the basis $\mathcal{C}$ as the matrix $\mathbf{C} = \begin{pmatrix} 0 & \frac{1}{2} \\ 1 & -\frac{1}{2} \end{pmatrix}$. Then, the change of basis can be represented as

```{r change-basis, cache = TRUE}
B <- matrix(c(1/2, 1, -1, 0), 2, 2)
C <- matrix(c(0, 1/2, 1, -1/2), 2, 2)

p <- plot_change_basis(B, C)
```


```{r change-basis-animation, cache = TRUE, message = FALSE, echo = FALSE, dependson = -1}
if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "html") {
  p  + gganimate::transition_states(time, wrap = FALSE)
}
```

which can be represented with the static images 

```{r change-basis-static, echo = FALSE}
p + facet_wrap(~ time, labeller = labeller(time = c("1" = "Basis B", "2" = "Basis C"))) 
```

The change of basis represents a linear transformation. When previously discussing linear transformations in Chapter \@ref(linear-transformations), we considered a linear transformation from the standard basis $\mathcal{I}$ defined by the basis vectors $\left\{ `r array_to_latex(c(1, 0))`, `r array_to_latex(c(0, 1))` \right\}$ with the vectors represented as the columns of the identity matrix $\mathbf{I}$. We can consider a change of basis as two consecutive linear transformations. First, a linear transformation from the basis $\mathcal{B}$ to the standard basis $\mathcal{I}$ and then a linear transformation from the standard basis $\mathcal{I}$ to the basis $\mathcal{C}$. This can be represented using the following example code:


```{r change-basis-I, cache = TRUE}
p <- plot_change_basis(B, C, plot_standard_basis = TRUE)
```


```{r change-basis-I-animation, cache = TRUE, message = FALSE, echo = FALSE, dependson = -1}
if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "html") {
  p  + gganimate::transition_states(time, wrap = FALSE)
}
```


```{r change-basis-I-static, echo = FALSE}
p + facet_wrap(~ time, labeller = labeller(time = c("1" = "Basis B", "1.5" = "Basis I", "2" = "Basis C"))) 
```


## Changing coordinates between different bases

Now, we can combine these ideas. Given a vector $\left[\mathbf{x}\right]_B$ written with coordinates with respect to the basis $\mathcal{B}$, we can find the coordinates of $\left[\mathbf{x}\right]_C$ with respect to the basis $\mathcal{C}$. First, we find the coordinates of the vector  $\mathbf{x}$ with respect to the standard basis then find the coordinates of $\left[\mathbf{x}\right]_C$ with respect to the basis $\mathcal{C}$. Let $\mathbf{B} = \begin{pmatrix} \mathbf{b}_1 & \ldots & \mathbf{b}_n \end{pmatrix}$ and $\mathbf{C} = \begin{pmatrix} \mathbf{c}_1 & \ldots & \mathbf{c}_n \end{pmatrix}$, then given a vector $\left[\mathbf{x}\right]_B$ with coordinates with respect to the basis $\mathcal{B}$, the coordinates $\left[\mathbf{x}\right]_C$ of this vector with respect to the basis $\mathcal{C}$ is 
$$
\begin{aligned}
\left[\mathbf{x}\right]_C = \mathbf{C}^{-1} \mathbf{B} \left[\mathbf{x}\right]_B.
\end{aligned}
$$

**Draw diagram**

::: {.callout-note icon=false collapse="true" appearance="simple"} 
## Solution

::: {#exm-}

```{r example-change-basis-1, echo = FALSE}
B <- matrix(c(1/2, 1, -1, 0), 2, 2)
C <- matrix(c(0, 1/2, 1, -1/2), 2, 2)
C_inv <- solve(C)
x_b <- c(-3/2, 1/2)
x <- B %*% x_b
x_c <- C_inv %*% x
```

Working with the same bases $\mathcal{B} = \left\{ \begin{pmatrix} \frac{1}{2} \\ 1 \end{pmatrix}, \begin{pmatrix} -1 \\ 0 \end{pmatrix}\right\}$ and $\mathcal{C} = \left\{ \begin{pmatrix} 0 \\ \frac{1}{2} \end{pmatrix}, \begin{pmatrix} 1 \\ -\frac{1}{2} \end{pmatrix}\right\}$ from the previous example, Let $\left[\mathbf{x}\right]_B = `r array_to_latex(x_b)`$ be the coordinates of the vector $\mathbf{x}$ with respect to the basis $\mathcal{B} = \left\{ `r array_to_latex(B[, 1])`, `r array_to_latex(B[, 2])`\right\}$. Find

1) the coordinates of $\mathbf{x}$ with respect to the standard basis and
2) the coordinates of $\mathbf{x}$ with respect to the basis $\mathcal{C}$.
:::


::: {.callout-note icon=false collapse="true" appearance="simple"} 
## Solution

Here we solve the two questions from the example above.

1) Recall that the coordinates $\left[\mathbf{x}\right]_B$ of $\mathbf{x}$ with respect to the basis $\mathcal{B}$ mean that the vector $\mathbf{x}$ can be written as a linear combination of the basis vectors $\mathbf{b}_1$ and $\mathbf{b}_2$ with coefficients given by the elements in $\left[\mathbf{x}\right]_B = \begin{pmatrix} \left[x_1\right]_B \\ \left[x_2\right]_B \end{pmatrix}$. This results in the equation

$$
\begin{aligned}
\mathbf{x} & = \left[x_1\right]_B \mathbf{b}_1 + \left[x_2\right]_B \mathbf{b}_2
\end{aligned}
$$
Plugging the values from the example gives 
$$
\begin{aligned}
\mathbf{x} & = \left[x_1\right]_B \mathbf{b}_1 + \left[x_2\right]_B \mathbf{b}_2 \\
& = `r MASS::fractions(x_b[1])` `r array_to_latex(B[, 1])` `r ifelse(x_b[2] >= 0, "+", "-")` `r MASS::fractions(abs(x_b[2]))` `r array_to_latex(B[, 2])` \\
& = `r array_to_latex(x_b[1] * B[, 1])` `r ifelse(x_b[2] >= 0, "+", "-")` `r array_to_latex(abs(x_b[2]) * B[, 2])` \\
& = `r array_to_latex(x)`
\end{aligned}
$$
2) Now, recall the coordinates $\left[\mathbf{x}\right]_C$ of $\mathbf{x}$ with respect to the basis $\mathcal{C}$ mean that the vector $\mathbf{x}$ can be written as a linear combination of the basis vectors $\mathbf{c}_1$ and $\mathbf{c}_2$ with coefficients given by the elements in $\left[\mathbf{x}\right]_C = \begin{pmatrix} \left[x_1\right]_C \\ \left[x_2\right]_C \end{pmatrix}$. This results in the equation

$$
\begin{aligned}
\mathbf{x} & = \left[x_1\right]_C \mathbf{c}_1 + \left[x_2\right]_C \mathbf{c}_2
\end{aligned}
$$
However, unlike part (1), we do not know the coefficients $\left[\mathbf{x}\right]$ but need to solve for them. Rewriting the above equation in the form of $\mathbf{A}\mathbf{x} = \mathbf{b}$ gives
$$
\begin{aligned}
\mathbf{C} \left[ \mathbf{x} \right]_C & = \mathbf{x}
\end{aligned}
$$
Because the matrix of basis vectors $\mathbf{C}$ is an invertible matrix (a basis is a linearly independent spanning set),  the coefficients $\left[\mathbf{x}\right]_C$ can be solved using the equation
$$
\begin{aligned}
\mathbf{C} \left[ \mathbf{x} \right]_C & = \mathbf{x} \\
\mathbf{C}^{-1} \mathbf{C}\left[ \mathbf{x} \right]_C & = \mathbf{C}^{-1}\mathbf{x} \\
\left[ \mathbf{x} \right]_C & = \mathbf{C}^{-1} \mathbf{x} 
\end{aligned}
$$
The matrix inverse $\mathbf{C}^{-1}$ can be found using @thm-matrix2by2 to get $\mathbf{C}^{-1} = `r array_to_latex(C_inv)`$. Then, plugging in the values from the example gives 

$$
\begin{aligned}
\left[ \mathbf{x} \right]_C & = \mathbf{C}^{-1} \mathbf{x} \\
& = `r array_to_latex(C_inv)` `r array_to_latex(x)` \\
& = `r array_to_latex(x_c)`
\end{aligned}
$$
Another way to change coordinates is to change directly from basis $\mathcal{B}$ to $\mathcal{C}$ without going through the intermediate transformation to the standard coordinates. Combining the results from (1) and (2) gives
$$
\begin{aligned}
\left[ \mathbf{x} \right]_C & = \mathbf{C}^{-1} \mathbf{x} \\
& = \mathbf{C}^{-1} \mathbf{B} \left[\mathbf{x}\right]_B
\end{aligned}
$$ 
so that one can change coordinates from the basis $\mathcal{B}$ to the basis $\mathcal{C}$ using the linear transformation defined by the matrix multiplication $\mathbf{C}^{-1} \mathbf{B}$.

In `R`, first define the basis matrices `B` and `C` and the coordinates `x_b` of the vector $\mathbf{x}$ with respect to the basis $\mathcal{B}$.

```{r example-change-basis-2}
B <- matrix(c(1/2, 1, -1, 0), 2, 2)
C <- matrix(c(0, 1/2, 1, -1/2), 2, 2)
x_b <- c(-3/2, 1/2)
```

1\) The coordinates `x` with respect to the standard basis is

```{r example-change-basis-3}
x <- B %*% x_b
x
```

2\) The coordinates `x_c` with respect to the basis $\mathcal{C}$ can be found by calculating the matrix inverse `C_inv` and then using the matrix inverse to calculate the coordinates with respect to the basis $\mathcal{C}$ as
```{r example-change-basis-4}
C_inv <- solve(C)
x_c <- C_inv %*% x
x_c
```

Done as a single transformation, the linear transformation is defined as

```{r example-change-basis-5}
B %*% C_inv
```
which gives the coordinates
```{r example-change-basis-6}
B %*% C_inv %*% x_b
```
:::


::: {#exm-}
3-d change of basis
:::




