# The Characteristic Equation


- [3 Blue 1 Brown -- ](https://www.3blue1brown.com/lessons/quick-eigen)


```{r, message = FALSE}
library(tidyverse)
library(dasc2594)
```

The characteristic equation/polynomial encodes information about the eigenvalues of the characteristic equation. In the previous chapter, we showed how we can decide if a scalar $\lambda$ is an eigenvalue of a matrix and how to find the vectors associated with the eigenvalue. However, we did not learn how to find eigenvalues (other than to just randomly try $\lambda$). The characteristic equation/polynomial allows for determining the eigenvalues $\lambda$.

:::{.definition}
Let $\mathbf{A}$ be a $n \times n$ matrix. The characteristic equation/polynomial of $\mathbf{A}$ is the function $f(\lambda)$ given by 
\begin{align*}
f(\lambda) = det(\mathbf{A} - \lambda \mathbf{I})
\end{align*}
:::

While not obvious, the function $f(\lambda)$ is a polynomial of $\lambda$ but requires computing the determinant of the matrix $\mathbf{A} - \lambda \mathbf{I}$ which contains an unknown value $\lambda$.


:::{.example}

Find the characteristic equation of the matrix $\mathbf{A} = \begin{pmatrix} 3 & 5 \\ 2 & -1 \end{pmatrix}$

* do in class
:::


:::{.example}

Find the characteristic equation of the matrix $\mathbf{A} = \begin{pmatrix} 0 & 6 & 8 \\ \frac{1}{2} & 0 & 0 \\ 0 & \frac{1}{2} & 0 \end{pmatrix}$

* do in class (expand cofactors along the third column)
:::

Once the characteristic equation is defined, we can use the equation to solve for the eigenvalues.

:::{.theorem}
Let $\mathbf{A}$ be a $n \times n$ matrix and let $f(\lambda) = det(\mathbf{A} - \lambda \mathbf{I})$ be a characteristic polynomial. Then, the number $\lambda_0$ is an eigenvalue of $\mathbf{A}$ if and only if $f(\lambda_0) = 0$.
:::


:::{.proof}
By the invertible matrix theorem, the matrix $(\mathbf{A} - \lambda_0 \mathbf{I}) \mathbf{x} = \mathbf{0}$ has a nontrivial solution if and only if $det(\mathbf{A} - \lambda_0 \mathbf{I}) \mathbf{x} = \mathbf{0}$. Therefore, the following statements are equivalent:

* $\lambda_0$ is an eigenvalue of $\mathbf{A}$ 

* $\mathbf{A} \mathbf{x} = \lambda_0 \mathbf{x}$ has a nontrivial solution

* $(\mathbf{A} - \lambda_0 \mathbf{I}) \mathbf{x} = \mathbf{0}$ has a nontrivial solution

* $\mathbf{A} - \lambda_0 \mathbf{I}$ is not invertible

* $det(\mathbf{A} - \lambda_0 \mathbf{I}) = \mathbf{0}$

* $f(\lambda_0) = 0$
:::

:::{.example}

Using the characteristic equation of the matrix $\mathbf{A} = \begin{pmatrix} 3 & 5 \\ 2 & -1 \end{pmatrix}$, solve for the eigenvalues and find a basis for the $\lambda$-eigenspaces

* do in class
:::


:::{.example}

Using the characteristic equation of the matrix $\mathbf{A} = \begin{pmatrix} 0 & 6 & 8 \\ \frac{1}{2} & 0 & 0 \\ 0 & \frac{1}{2} & 0 \end{pmatrix}$, solve for the eigenvalues and find a basis for the $\lambda$-eigenspaces

* do in class (expand cofactors along the third column)
:::


## Similarity

The idea behind similar matrices is to understand how the linear transformations implied by the transformation behave. Two matrices are similar if their transformation behavior (rotation, expansion/contraction, etc.) is the same but the coordinates on which the matrix operates are different. 



:::{.definition}
The matrices $\mathbf{A}$ and $\mathbf{B}$ are said to be similar if there exists an invertible matrix $\mathbf{P}$ where 
\begin{align*}
\mathbf{A} = \mathbf{P} \mathbf{B} \mathbf{P}^{-1}
\end{align*}
or equivalently
\begin{align*}
\mathbf{P}^{-1} \mathbf{A} \mathbf{P}=  \mathbf{B} 
\end{align*}
:::

Therefore, it is possible to change $\mathbf{A}$ into $\mathbf{B}$ with an invertible (one-to-one and onto) transformation. 

:::{.example}
Consider the following example with matrices $\mathbf{A}$, $\mathbf{B}$, and $\mathbf{P}$ defined as below:
```{r}
A <- matrix(c(3, 0, 0, -2), 2, 2)
A

B <- matrix(c(-12, -10, 15, 13), 2, 2)
B

P <- matrix(c(-2, 1, 3,-1), 2, 2)
P

P %*% B %*% solve(P)
solve(P) %*% A %*% P
```
:::

:::{.theorem}
If $\mathbf{A}$ and $\mathbf{B}$ are $n \times n$ similar matrices, then $\mathbf{A}$ and $\mathbf{B}$ will have the same characteristic polynomial and therefore the same eigenvalues.
:::


:::{.proof}
If $\mathbf{A}$ and $\mathbf{B}$ are similar, then there exists an invertible matrix $\mathbf{P}$ such that
\begin{align*}
\mathbf{A} = \mathbf{P} \mathbf{B} \mathbf{P}^{-1}
\end{align*}
Therefore
\begin{align*}
\mathbf{A}  - \lambda \mathbf{I} & = \mathbf{P} \mathbf{B} \mathbf{P}^{-1} - \lambda \mathbf{I} \\
& = \mathbf{P} \mathbf{B} \mathbf{P}^{-1} - \lambda \mathbf{P} \mathbf{P}^{-1} \\
& =  \mathbf{P} \left( \mathbf{B} \mathbf{P}^{-1} - \lambda \mathbf{P}^{-1} \right) \\
& =  \mathbf{P} \left( \mathbf{B} - \lambda \mathbf{I} \right) \mathbf{P}^{-1}\\
\end{align*}
To get the characteristic equation, we need to solve for the determinant 
\begin{align*}
det\left( \mathbf{A}  - \lambda \mathbf{I} \right) & = det\left( \mathbf{P} \left( \mathbf{B} - \lambda \mathbf{I} \right) \mathbf{P}^{-1} \right) \\
& = det\left( \mathbf{P} \right)  det\left( \mathbf{B} - \lambda \mathbf{I} \right) det\left(\mathbf{P}^{-1} \right) \\
\end{align*}
We know that $det\left(\mathbf{P}^{-1} \right)$ = $\frac{1}{det\left(\mathbf{P} \right)}$ (or, equivalently $det\left(\mathbf{P} \right) det\left(\mathbf{P}^{-1} \right) = det\left(\mathbf{P} \mathbf{P}^{-1} \right) = det(\mathbf{I}) = 1$), we have $det\left( \mathbf{A}  - \lambda \mathbf{I} \right) = det\left( \mathbf{B} - \lambda \mathbf{I} \right)$ so that $\mathbf{A}$ and $\mathbf{B}$ have the same characteristic polynomial (and the same eigenvalues).
:::

## The geometric interpetation of similar matrices

In general, similar matrices do similar things in different spaces (different spaces in terms of different bases).

[Example here](https://textbooks.math.gatech.edu/ila/similarity.html)






