# 


```{r, message = FALSE}
library(tidyverse)
library(dasc2594)
library(rgl)
set.seed(2021)
```

Here we start a transition to topics in vector calculus. We will start with a discussion of functions of two variables (although the functions are not assumed to be linear). We define a function of two variables explicitly as $z = f(x, y)$. 

Domain and range



### Graphs and level curves

:::{.example}
The parabola

Consider the function of two variables 
\begin{align*}
f(x, y) = x^2 + y^2
\end{align*}
which defines the surface
```{r, echo = FALSE}
parabola <- function(x, y) {
    return(x^2 + y^2)
}
# generate a grid over which to plot
x <- y <- seq(-1, 1, length = 50)
# apply the function to the grid
z <- outer(x, y, parabola)
# generate the plot
persp(x, y, z)
```
The 3-dimensional surface can be represented in 2-dimensions using level curves (think of a topographic map)
```{r}
data.frame(expand.grid(x, y)) %>%
    rename(x = Var1, y = Var2) %>%
    mutate(z = parabola(x, y)) %>%
    ggplot(aes(x = x, y = y, z = z)) +
    geom_contour() +
    coord_fixed(ratio = 1)
```
where each curve in the (x, y) plane has exactly the same value of $f(x, y)$. Alternatively, this can be represented using filled level curves 
```{r}
data.frame(expand.grid(x, y)) %>%
    rename(x = Var1, y = Var2) %>%
    mutate(z = parabola(x, y)) %>%
    ggplot(aes(x = x, y = y, z = z)) +
    geom_contour_filled() +
    coord_fixed(ratio = 1)
```
Notice that although the original parabola was continuous, these 2-d representations simplify the diagram by representing the contours as discrete values. 
:::


:::{.example}
A saddle function

Consider the function of two variables 
\begin{align*}
f(x, y) = x^2 - y^2
\end{align*}
which defines the surface
```{r, echo = FALSE}
saddle <- function(x, y) {
    return(x^2 - y^2)
}
# generate a grid over which to plot
x <- y <- seq(-1, 1, length = 50)
# apply the function to the grid
z <- outer(x, y, saddle)
# generate the plot
persp(x, y, z, theta = 33.5)
```
The 3-dimensional surface can be represented in 2-dimensions using level curves (think of a topographic map)
```{r}
data.frame(expand.grid(x, y)) %>%
    rename(x = Var1, y = Var2) %>%
    mutate(z = saddle(x, y)) %>%
    ggplot(aes(x = x, y = y, z = z)) +
    geom_contour() +
    coord_fixed(ratio = 1)
```
where each curve in the (x, y) plane has exactly the same value of $f(x, y)$. Alternatively, this can be represented using filled level curves 
```{r}
data.frame(expand.grid(x, y)) %>%
    rename(x = Var1, y = Var2) %>%
    mutate(z = saddle(x, y)) %>%
    ggplot(aes(x = x, y = y, z = z)) +
    geom_contour_filled() +
    coord_fixed(ratio = 1)
```
Notice that although the original saddle was continuous, these 2-d representations simplify the diagram by representing the contours as discrete values. 
:::




### Limits and continuity
For functions of several variables, we have to define limits and continuity for these multivariable settings. For now, we focus on two variable functions as the multivariable case follows similar from the two variable case.

Let $P(x, y) \rightarrow P_0(a, b)$ be a path in the $x-y$ plane that starts at the point $P(x, y)$ and ends at the point $P_0(a, b)$ with coordinates $(a, b)$. Thus, we can understand the limit as the fixed value of $f(x, y)$ for which all paths that connect the points $P(x, y)$ that are "close" to $P_0(a, b)$ converge to. 

For one-dimensional limits, "close" was defined as distance. Thus, for multivariable functions, "close" is defined as the Euclidean distance defined by a "ball" of radius $\delta$ and the limits examines the function output as the radius $\delta$ goes to 0.

Recall that the distance $dist((x, y), (a, b))$ between two points $(x, y)$ and $(a, b)$ is 
\begin{align*}
dist((x, y), (a, b)) = \sqrt{(x-a)^2 + (y-b)^2}
\end{align*}

** Draw images**


```{definition, name = "Limit of a function of two variables"}
The function f(x, y) has limit $L$ as $P(x, y) \rightarrow P_0(a, b)$, written
\begin{align*}
\lim_{(x, y) \rightarrow (a, b)} f(x, y) = \lim_{P \rightarrow P_0} f(x, y) = L,
\end{align*}
if, for any $\epsilon > 0$ (the radius of the ball that defines the "closeness" of the point), there exists a $\delta > 0$ such that 
\begin{align*}
|f(x, y) - L| < \epsilon
\end{align*}
whenever $(x, y)$ is in the domain of $f$ and 
\begin{align*}
0 < \sqrt{(x-a)^2 + (y-b)^2} < \delta.
\end{align*}
```

In the definition, as the value of $\delta$ is getting smaller, the distance between the set of points $P(x, y)$ within radius $\delta$ of the point $P_0(a, b)$ is getting smaller. As a consequence, the limit in the definition above exists only if $f(x, y)$ approaches the value $L$ along **all possible paths** in the domain of $f$. 






