--- 
title: "Multivariable Mathematics for Data Science"
author: "John Tipton"
date: "`r Sys.Date()`"
output: pdf_document
description: This is a minimal example of using the bookdown package to write a book.
  The output format for this example is bookdown::gitbook.
documentclass: book
link-citations: yes
bibliography:
- book.bib
- packages.bib
site: bookdown::bookdown_site
runtime: shiny
biblio-style: apalike
---

```{r include = FALSE, echo = FALSE, message = FALSE}
library(tidyverse)
library(shiny)
```

# Preface

This book will introduce students to multivariable Calculus and linear algebra methods and techniques to be successful in data science, statistics, computer science, and other data-driven, computational disciplines. 

The motiviation for this text is to provide both a theoretical understanding of important multivariable methods used in data science as well as giving a hands-on experience using software. Throughout this text, we assume the reader has a solid foundation in univariate calculus (typically two semesters) as well as familiarity with a scripting language (e.g., R or python).



## Outline

1) Introduction to vectors and matrices and vector and matrix operations

1) vector spaces and subspaces

2) dot products, cross products, projections

3) linear combinations, linear independence, bases, coordinate systems

4) planes, surfaces, and lines in space

5) linear transformations, matrix arithmetic, and matrix rank

6) solving linear equations $\mathbf{A} \mathbf{x} = \mathbf{b}$

7) innner products, outer products, and norms

8) projections - orthogonal projections and least squares

9) matrix decompositoins: Eigen, Cholesky, principal components, 
singular value decomposition, pre-whitening

10) matrix decompositoins: Eigen, Cholesky, principal components, 
singular value decomposition, pre-whitening

11) limits, continuity, and partial derivatives

12) chain rule, directional derivatives

13) tangent planes, linear approximations, Taylor Series

14) double/triple integrals, change of variables, Jacobian transformation

15) double/triple integrals, change of variables, Jacobian transformation

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

