--- 
title: "Multivariable Mathematics for Data Science"
author: "John Tipton"
date: "`r Sys.Date()`"
output: pdf_document
description: This is a minimal example of using the bookdown package to write a book.
  The output format for this example is bookdown::gitbook.
documentclass: book
link-citations: yes
bibliography:
- book.bib
- packages.bib
site: bookdown::bookdown_site
runtime: shiny
biblio-style: apalike
header-includes: preamble.tex
---

```{r include = FALSE, echo = FALSE, message = FALSE}
library(tidyverse)
library(shiny)

# if devtools package not installed, install the package
if (!require(devtools)) {
    install.packages("devtools")
}
# if dasc2594 package not installed, install the package
if (!require(dasc2594)) {
    devtools::install_github("jtipton25/dasc2594")
}
```

# Preface

This book will introduce students to multivariable Calculus and linear algebra methods and techniques to be successful in data science, statistics, computer science, and other data-driven, computational disciplines. 

The motiviation for this text is to provide both a theoretical understanding of important multivariable methods used in data science as well as giving a hands-on experience using software. Throughout this text, we assume the reader has a solid foundation in univariate calculus (typically two semesters) as well as familiarity with a scripting language (e.g., R or python).



## Getting started in `R`

## Matrices

We let uppercase bold letters $\mathbf{A}$, $\mathbf{B}$, etc., represent matrices. We define the matrix $\mathbf{A}$ with $m$ rows and $n$ columns as

$$
\begin{align*}
\mathbf{A} & = \begin{pmatrix} a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{pmatrix},
\end{align*}
$$

with $a_{ij}$ being the value of the matrix $\mathbf{A}$ in the $i$th row and the $j$th column.

If the matrix 

$$
\begin{align*}
\mathbf{A} & = \begin{pmatrix} 5 & 7 & 1 \\
5 & -22  & 2 \\
-14 & 5 & 99 \\
42 & -3 & 0\end{pmatrix},
\end{align*}
$$

the elements $a_{11}$ = 5, $a_{12}$ = 7, $a_{21}$ = 5, and $a_{33}$ = 99, etc. 

In `R`, we can define the matrix A using the `matrix()` function

```{r}
A <- matrix(
    data = c(5, 5, -14, 42, 7, -22, 5, -3, 1, 2, 99, 0),
    nrow = 4,
    ncol = 3
)

A
```

Notice in the above creation of $\mathbf{A}$, we wrote defined the elements of the  $\mathbf{A}$ using the columns stacked on top of one another. If we want to fill in the elements of $\mathbf{A}$ using the rows, we can add the option `byrow = TRUE` to the `matrix()` function

```{r}
A <- matrix(
    data  = c(5, 7, 1, 5, -22, 2, -14, 5, 99, 42, -3, 0), 
    nrow  = 4,
    ncol  = 3,
    byrow = TRUE
)
A
```

To select the $ij$th elements of $\mathbf{A}$, we use the subset operator `[` to select the element. For example, to get the element $a_{11} = 5$ in the first row and first column of $\mathbf{A}$, we use

```{r}
A[1, 1]
```

The element $a_{3, 3} = 99$ in the third row and third column can be selected using

```{r}
A[3, 3]
```


The matrix $\mathbf{A}$ can also be represented as a set of either column vectors $\{\mathbf{c}_j \}_{j=1}^p$ or row vectors $\{\mathbf{r}_i \}_{i=1}^n$. For example, the column vector representation is


$$
\begin{align*}
\mathbf{A} & = \left( \mathbf{c}_{1} \middle| \mathbf{c}_{2} \middle| \cdots \middle| \mathbf{c}_{p} \right),
\end{align*}
$$

where the notation $|$ is used to separate the vectors

$$
\begin{align*}
\mathbf{c}_1 & = \begin{pmatrix} a_{11} \\ a_{21} \\ \vdots \\ a_{n1}
\end{pmatrix},
& \mathbf{c}_2 & = \begin{pmatrix} a_{12} \\ a_{22} \\ \vdots \\ a_{n2}
\end{pmatrix}, 
& \cdots, &
& \mathbf{c}_p & = \begin{pmatrix} a_{1p} \\ a_{2p} \\ \vdots \\ a_{np}
\end{pmatrix}
\end{align*}
$$

In `R` you can extract the columns using the `[` selection operator

```{r}
c1 <- A[, 1] # first column
c2 <- A[, 2] # second column
c3 <- A[, 3] # third column
```

and you can give the column representation of the matrix `A` with with column bind function `cbind()`

```{r}
cbind(c1, c2, c3)
```

The row vector representation of $\mathbf{A}$ is 

$$
\begin{align*}
\mathbf{A} & = \begin{pmatrix} \mathbf{r}_{1} \\ \mathbf{r}_{2} \\ \vdots \\ \mathbf{r}_{n} \end{pmatrix},
\end{align*}
$$

where the row vectors $\mathbf{r}_i$ are

$$
\begin{align*}
\mathbf{r}_1 & = \left( a_{11}, a_{12}, \ldots, a_{1n} \right) \\
\mathbf{r}_2 & = \left( a_{21}, a_{22}, \ldots, a_{2n} \right) \\
& \vdots \\
\mathbf{r}_m & = \left( a_{m1}, a_{m2}, \ldots, a_{mn} \right)
\end{align*}
$$


In `R` you can extract the rows using the `[` selection operator

```{r}
r1 <- A[1, ] # first row
r2 <- A[2, ] # second row
r3 <- A[3, ] # third row
r4 <- A[4, ] # fourth row
```

and you can give the row representation of the matrix `A` with with row bind function `rbind()`

```{r}
rbind(r1, r2, r3, r4)
```




<!-- ## Outline -->

<!-- **Note: This is going to be revised and reformated throughout the class** -->

<!-- 1) Introduction to vectors and matrices and vector and matrix operations -->

<!-- 1) vector spaces and subspaces -->

<!-- 2) dot products, cross products, projections -->

<!-- 3) linear combinations, linear independence, bases, coordinate systems -->

<!-- 4) planes, surfaces, and lines in space -->

<!-- 5) linear transformations, matrix arithmetic, and matrix rank -->

<!-- 6) solving linear equations $\mathbf{A} \mathbf{x} = \mathbf{b}$ -->

<!-- 7) innner products, outer products, and norms -->

<!-- 8) projections - orthogonal projections and least squares -->

<!-- 9) matrix decompositoins: Eigen, Cholesky, principal components,  -->
<!-- singular value decomposition, pre-whitening -->

<!-- 10) matrix decompositoins: Eigen, Cholesky, principal components,  -->
<!-- singular value decomposition, pre-whitening -->

<!-- 11) limits, continuity, and partial derivatives -->

<!-- 12) chain rule, directional derivatives -->

<!-- 13) tangent planes, linear approximations, Taylor Series -->

<!-- 14) double/triple integrals, change of variables, Jacobian transformation -->

<!-- 15) double/triple integrals, change of variables, Jacobian transformation -->

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```



<!-- Notes for revision -->
<!-- 1) Start with vector and matrix addition/multiplication properties. Show how to  -->
<!-- 2) Introduce how to solve a matrix equation with `solve()`. Notice that sometimes this works and sometimes this fails. Maybe we can understand why -->
<!-- 3) Introduce systems of equations and rref  -->
<!-- 4) Show how Gaussian-Jordan elimination can solve these equations using augmented matrices -->
<!-- 5) Introduce linear combinations, spans, linear independence -->
<!-- 6) Show how the concepts in 5) determine if a solution exists (invertibility) -->

